---
title: "Getting started: connecting and selecting locations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: connecting and selecting locations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(watina)
library(sf)
library(dplyr)
library(stringr)
library(knitr)
```

_General note: the below vignette contains frozen output of 16 Sep 2019._
_This makes it possible to build the package with vignettes without access to the Watina database._

## Connecting

You first need a connection object to connect to Watina; let's simply name it `watina`:

```{r}
watina <- connect_watina()
```


## Selecting locations

Selecting locations is a typical starting point.
Afterwards, you can use the returned object as input to other functions.

By default, a `tbl_lazy` object is given (lazy query).^[
See the [dbplyr](https://dbplyr.tidyverse.org/) package to get you started with this type of object.
]
The returned data include some essential metadata of the locations.
The following example shows a few rows:

```{r}
get_locs(watina) %>% 
  head(5) %>% 
  collect %>% 
  as.data.frame
#>   loc_wid loc_code area_code area_name      x      y loc_validitycode
#> 1     676  AABP011       AAB    Aabeek 178331 209751              VLD
#> 2     677  AABP013       AAB    Aabeek 178315 209637              VLD
#> 3     678  AABP014       AAB    Aabeek 178310 209590              VLD
#> 4     681  AABP017       AAB    Aabeek 178370 209461              VLD
#> 5     682  AABP018       AAB    Aabeek 178379 209418              VLD
#>   loc_validity loc_typecode        loc_typename
#> 1  Gevalideerd            P Peilbuis/Piëzometer
#> 2  Gevalideerd            P Peilbuis/Piëzometer
#> 3  Gevalideerd            P Peilbuis/Piëzometer
#> 4  Gevalideerd            P Peilbuis/Piëzometer
#> 5  Gevalideerd            P Peilbuis/Piëzometer
```

Selections can be done in a variety of ways -- and these ways can be combined:

- by using area codes:

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"))
#> # Source:     lazy query [?? x 10]
#> # Database:   Microsoft SQL Server
#> # Ordered by: area_code, loc_code
#>    loc_wid loc_code area_code area_name      x      y loc_validitycode
#>      <int> <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>           
#>  1    2824 KALP022  KAL       Kalmthou… 155350 230813 VLD             
#>  2    1546 KALP030  KAL       Kalmthou… 152990 238600 VLD             
#>  3   10069 KALP034  KAL       Kalmthou… 152962 231345 VLD             
#>  4    1549 KALP035  KAL       Kalmthou… 153000 236850 VLD             
#>  5    1552 KALP039  KAL       Kalmthou… 160940 235060 VLD             
#>  6    4327 KALP050  KAL       Kalmthou… 153881 233802 VLD             
#>  7    6210 KALP051  KAL       Kalmthou… 153822 233761 VLD             
#>  8    1562 KALP052  KAL       Kalmthou… 153822 233761 VLD             
#>  9    1563 KALP053  KAL       Kalmthou… 153781 233736 VLD             
#> 10    6180 KALP054  KAL       Kalmthou… 154689 232426 VLD             
#> # … with more rows, and 3 more variables: loc_validity <chr>,
#> #   loc_typecode <chr>, loc_typename <chr>
```

The number of involved locations is given by:

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR")) %>% 
  count
#> # Source:   lazy query [?? x 1]
#> # Database: Microsoft SQL Server
#>       n
#>   <int>
#> 1   188
```


- you can specify the type of location (device):

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"),
         loc_type = c("P", "S")) %>% 
  count
#> # Source:   lazy query [?? x 1]
#> # Database: Microsoft SQL Server
#>       n
#>   <int>
#> 1   223
```

- by using a spatial bounding box (rectangular area), using the Belgian Lambert 72 coordinate reference system (EPSG-code [31370](https://epsg.io/31370)):

```{r}
get_locs(watina,
         bbox = c(xmin = 1.4e+5,
                  xmax = 1.7e+5,
                  ymin = 1.6e+5,
                  ymax = 1.9e+5))
#> # Source:     lazy query [?? x 10]
#> # Database:   Microsoft SQL Server
#> # Ordered by: area_code, loc_code
#>    loc_wid loc_code area_code area_name      x      y loc_validitycode
#>      <int> <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>           
#>  1   10261 ATGP002  ATG       Antitank… 169572 186043 VLD             
#>  2    5629 DORP016  DOR       Dorent    158521 186074 VLD             
#>  3    5630 DORP017  DOR       Dorent    158442 185925 VLD             
#>  4    5632 DORP019  DOR       Dorent    158304 184915 VLD             
#>  5    5634 DORP020  DOR       Dorent    158352 184259 VLD             
#>  6    5635 DORP021  DOR       Dorent    158103 183853 VLD             
#>  7    5586 DORP022  DOR       Dorent    157715 184013 VLD             
#>  8    5587 DORP023  DOR       Dorent    154284 182533 VLD             
#>  9    6420 DORP024  DOR       Dorent    154823 182431 VLD             
#> 10    5580 DORP027  DOR       Dorent    155151 182190 VLD             
#> # … with more rows, and 3 more variables: loc_validity <chr>,
#> #   loc_typecode <chr>, loc_typename <chr>

```

- if you already have specific location codes in mind, you can provide these directly as a vector:

```{r}
get_locs(watina,
         loc_vec = c("KBRP081", "KBRP090", "KBRP095"))
#> # Source:     lazy query [?? x 10]
#> # Database:   Microsoft SQL Server
#> # Ordered by: area_code, loc_code
#>   loc_wid loc_code area_code area_name      x      y loc_validitycode
#>     <int> <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>           
#> 1     664 KBRP081  KBR       Potpolde… 145846 204871 VLD             
#> 2     698 KBRP090  KBR       Potpolde… 145988 204702 VLD             
#> 3     709 KBRP095  KBR       Potpolde… 145949 204712 VLD             
#> # … with 3 more variables: loc_validity <chr>, loc_typecode <chr>,
#> #   loc_typename <chr>

```

- you can use a geospatial filter of class [sf](https://r-spatial.github.io/sf/) to make selections, by using the `mask` argument:

```{r echo=FALSE, include=FALSE}
data(BE_ADMIN_PROVINCE,
     package = "BelgiumMaps.StatBel")
westfl <-
    st_as_sf(BE_ADMIN_PROVINCE) %>%
    filter(str_detect(TX_PROV_DESCR_NL, "West")) %>%
    st_transform(crs = 31370)
```


```{r}
get_locs(watina,
         mask = westfl,
         buffer = 0)
#> As a mask always invokes a collect(), the argument 'collect = FALSE' will be ignored.
#> Warning in get_locs(watina, mask = westfl, buffer = 0): Dropped 34
#> locations from which x or y coordinates were missing.
#> # A tibble: 740 x 9
#>    loc_code area_code area_name     x      y loc_validitycode loc_validity
#>  * <chr>    <chr>     <chr>     <dbl>  <dbl> <chr>            <chr>       
#>  1 BELP009  BEL       de Belvé… 28145 199761 VLD              Gevalideerd 
#>  2 BELP010  BEL       de Belvé… 27947 199669 VLD              Gevalideerd 
#>  3 OSKP043  OSK       Oostends… 54040 210979 VLD              Gevalideerd 
#>  4 DHEP002  DHE       D' Heye   53635 216178 VLD              Gevalideerd 
#>  5 DHEP003  DHE       D' Heye   53596 215858 VLD              Gevalideerd 
#>  6 DHEP004  DHE       D' Heye   53841 216470 VLD              Gevalideerd 
#>  7 ZEBP032  ZEB       Zeebrugge 72142 218026 VLD              Gevalideerd 
#>  8 KEMP002  KEM       Kemmelbe… 39912 163431 VLD              Gevalideerd 
#>  9 KEMP003  KEM       Kemmelbe… 40561 163750 VLD              Gevalideerd 
#> 10 KEMP005  KEM       Kemmelbe… 39947 164778 VLD              Gevalideerd 
#> # … with 730 more rows, and 2 more variables: loc_typecode <chr>,
#> #   loc_typename <chr>
```

Note that the default value for `buffer` is 10 meters; it is used to enlarge mask (or shrink it, if `buffer` < 0).

By default, only locations are returned which have been validated, i.e. with `loc_validity = "VLD"`.
This can however be relaxed, for example by also including locations that are not yet validated:

```{r}
get_locs(watina,
         area_codes = c("ZWA"),
         loc_type = c("P", "S"),
         loc_validity = c("VLD", "ENT")
         ) %>% 
  count
#> # Source:   lazy query [?? x 1]
#> # Database: Microsoft SQL Server
#>       n
#>   <int>
#> 1   222
```

Instead of leaving the output as a `tbl_lazy` object, you can also retrieve it as a tibble (i.e. an easier-to-use type of dataframe) by setting `collect = TRUE`.

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"),
         loc_type = c("P", "S"),
         loc_validity = c("VLD", "ENT"),
         collect = TRUE
         )
#> # A tibble: 223 x 9
#>    loc_code area_code area_name      x      y loc_validitycode loc_validity
#>    <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>            <chr>       
#>  1 KBRP081  KBR       Potpolde… 145846 204871 VLD              Gevalideerd 
#>  2 KBRP090  KBR       Potpolde… 145988 204702 VLD              Gevalideerd 
#>  3 KBRP095  KBR       Potpolde… 145949 204712 VLD              Gevalideerd 
#>  4 KBRS001  KBR       Potpolde… 146937 205586 VLD              Gevalideerd 
#>  5 KALP030  KAL       Kalmthou… 152990 238600 VLD              Gevalideerd 
#>  6 KALP035  KAL       Kalmthou… 153000 236850 VLD              Gevalideerd 
#>  7 KALP039  KAL       Kalmthou… 160940 235060 VLD              Gevalideerd 
#>  8 KALP052  KAL       Kalmthou… 153822 233761 VLD              Gevalideerd 
#>  9 KALP053  KAL       Kalmthou… 153781 233736 VLD              Gevalideerd 
#> 10 KALP059  KAL       Kalmthou… 154754 232459 VLD              Gevalideerd 
#> # … with 213 more rows, and 2 more variables: loc_typecode <chr>,
#> #   loc_typename <chr>
```

Note that the lazy object contains a `loc_wid` column, which is an internal database variable to identify locations and make relations between tables.
It should not be regarded as a stable ID; hence don't store it for later use (note that lazy objects don't store data).
Consequently, the `loc_wid` column is omitted when using `collect = TRUE`.

## Disconnecting

At the end of your work (a script, a bookdown document etc.), close the connection:

```{r}
DBI::dbDisconnect(watina)
```



