---
title: "Getting started: connecting and selecting locations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connecting and retrieving locations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(watina)
library(sf)
library(dplyr)
library(stringr)
library(knitr)
```

## Connecting

You first need a connection object to connect to Watina; let's simply name it `watina`:

```{r}
watina <- connect_watina()
```


## Selecting locations

Selecting locations is a typical starting point.
Afterwards, you can use the returned object as input for other functions.

By default, a `tbl_lazy` object is given (lazy query).^[
See the [dbplyr](https://dbplyr.tidyverse.org/) package to get you started with that type of object.
]
The returned data include some essential metadata of the locations.
The following example shows a few rows:

```{r}
get_locs(watina) %>% 
  head(5) %>% 
  collect %>% 
  kable
```

Selections can be done in a variety of ways -- they can be combined:

- by using area codes:

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"))
```

The number of involved locations is given by:

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR")) %>% 
  count
```


- you can specify the type of location (device):

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"),
         loc_type = c("P", "S")) %>% 
  count
```

- by using a spatial bounding box (rectangular area), using the Belgian Lambert 72 coordinate reference system (EPSG-code [31370](https://epsg.io/31370)):

```{r}
get_locs(watina,
         bbox = c(xmin = 1.4e+5,
                  xmax = 1.7e+5,
                  ymin = 1.6e+5,
                  ymax = 1.9e+5))
```

- if you already have specific location codes in mind, you can provide these directly as a vector:

```{r}
get_locs(watina,
         loc_vec = c("KBRP081", "KBRP090", "KBRP095"))
```

- you can use a geospatial filter of class [sf](https://r-spatial.github.io/sf/) to make selections, by using the `mask` argument:

```{r echo=FALSE, include=FALSE}
data(BE_ADMIN_PROVINCE,
     package = "BelgiumMaps.StatBel")
westfl <-
    st_as_sf(BE_ADMIN_PROVINCE) %>%
    filter(str_detect(TX_PROV_DESCR_NL, "West")) %>%
    st_transform(crs = 31370)
```


```{r}
get_locs(watina,
         mask = westfl,
         buffer = 0)
```

Note that the default value for `buffer` is 10 meters; it is used to enlarge mask (or shrink it, if `buffer` < 0).

By default, only locations are returned which have been validated, i.e. with `loc_validity = "VLD"`.
This can however be relaxed, for example by also including locations that are not yet validated:

```{r}
get_locs(watina,
         area_codes = c("ZWA"),
         loc_type = c("P", "S"),
         loc_validity = c("VLD", "ENT")
         ) %>% 
  count
```

Instead of leaving the output as a `tbl_lazy` object, you can also retrieve it as a tibble (i.e. an easier-to-use type of dataframe) by setting `collect = TRUE`.

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"),
         loc_type = c("P", "S"),
         loc_validity = c("VLD", "ENT"),
         collect = TRUE
         )
```

Note that the lazy object contains a `loc_wid` column, which is an internal database variable to identify locations and make relations between tables.
It should not be regarded as a stable ID; hence don't store it for later use (note that lazy objects don't store data).
Consequently, the `loc_wid` column is omitted when using `collect = TRUE`.

## Disconnecting

At the end of your work (a script, a bookdown document etc.), close the connection:

```{r}
DBI::dbDisconnect(watina)
```



