---
title: "Getting started: connecting and selecting locations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: connecting and selecting locations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(watina)
library(sf)
library(dplyr)
library(stringr)
library(knitr)
```

_General note: the below vignette contains frozen output of 17 Sep 2019._
_This makes it possible to build the package with vignettes without access to the Watina database._

## Connecting

You first need a connection object to connect to Watina; let's simply name it `watina`:

```{r}
watina <- connect_watina()
```


## Selecting locations

Selecting locations is a typical starting point.
Afterwards, you can use the returned object as input to other functions.

By default, a `tbl_lazy` object is given (lazy query).^[
See the [dbplyr](https://dbplyr.tidyverse.org/) package to get you started with this type of object.
]
The returned data include some essential metadata of the locations.
The following example shows a few rows:

```{r}
get_locs(watina) %>% 
  head(5) %>% 
  collect %>% 
  as.data.frame
#>   loc_wid loc_code area_code area_name      x      y loc_validitycode
#> 1     676  AABP011       AAB    Aabeek 178331 209751              VLD
#> 2     677  AABP013       AAB    Aabeek 178315 209637              VLD
#> 3     678  AABP014       AAB    Aabeek 178310 209590              VLD
#> 4     681  AABP017       AAB    Aabeek 178370 209461              VLD
#> 5     682  AABP018       AAB    Aabeek 178379 209418              VLD
#>   loc_validity loc_typecode        loc_typename filterdepth soilsurf_ost
#> 1  Gevalideerd            P Peilbuis/Piëzometer        2.90        10.34
#> 2  Gevalideerd            P Peilbuis/Piëzometer        2.71        10.34
#> 3  Gevalideerd            P Peilbuis/Piëzometer        2.74         9.41
#> 4  Gevalideerd            P Peilbuis/Piëzometer        2.72         9.31
#> 5  Gevalideerd            P Peilbuis/Piëzometer        2.58         9.50
```

Selections can be done in a variety of ways -- and these ways can be combined:

- by using area codes:

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"))
#> # Source:     lazy query [?? x 12]
#> # Database:   Microsoft SQL Server
#> # Ordered by: area_code, loc_code, obswell_rank, area_code, loc_code
#>    loc_wid loc_code area_code area_name      x      y loc_validitycode
#>      <int> <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>           
#>  1    2824 KALP022  KAL       Kalmthou… 155350 230813 VLD             
#>  2    1546 KALP030  KAL       Kalmthou… 152990 238600 VLD             
#>  3   10069 KALP034  KAL       Kalmthou… 152962 231345 VLD             
#>  4    1549 KALP035  KAL       Kalmthou… 153000 236850 VLD             
#>  5    1552 KALP039  KAL       Kalmthou… 160940 235060 VLD             
#>  6    4327 KALP050  KAL       Kalmthou… 153881 233802 VLD             
#>  7    6210 KALP051  KAL       Kalmthou… 153822 233761 VLD             
#>  8    1562 KALP052  KAL       Kalmthou… 153822 233761 VLD             
#>  9    1563 KALP053  KAL       Kalmthou… 153781 233736 VLD             
#> 10    6180 KALP054  KAL       Kalmthou… 154689 232426 VLD             
#> # … with more rows, and 5 more variables: loc_validity <chr>,
#> #   loc_typecode <chr>, loc_typename <chr>, filterdepth <dbl>,
#> #   soilsurf_ost <dbl>
```

The number of involved locations is given by:

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR")) %>% 
  count
#> # Source:   lazy query [?? x 1]
#> # Database: Microsoft SQL Server
#>       n
#>   <int>
#> 1   188
```


- you can specify the type of location (device):

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"),
         loc_type = c("P", "S")) %>% 
  count
#> # Source:   lazy query [?? x 1]
#> # Database: Microsoft SQL Server
#>       n
#>   <int>
#> 1   223
```

- for groundwater piezometers: by setting conditions on filterdepth.
The filterdepth is the depth of the piezometer's filter below soil surface (unit: m), more specifically referring to _halfway_ the length of the filter.
Conditions can be set by the arguments `filterdepth_range`  and `filterdepth_na` (see documentation of `get_locs()`).
Moreover, by setting `filterdepth_guess = TRUE` (default `FALSE`) you can replace missing filterdepths by a conservative value, before applying these conditions.
By default, only filterdepths between 0 and 3 meters below soilsurface are returned and locations with missing filterdepth are dropped.

Indeed, more records are returned when allowing groundwater piezometers with missing filterdepth:

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"),
         loc_type = c("P", "S"),
         filterdepth_na = TRUE) %>% 
  count
```

Of course you can set further conditions for other variables, by building upon the lazy result of `get_locs()`, e.g.:

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"),
         loc_type = c("P", "S"),
         filterdepth_na = TRUE) %>%
  filter(obswell_state == "OK - niet bemeten") %>% 
  count
```


- by using a spatial bounding box (rectangular area), using the Belgian Lambert 72 coordinate reference system (EPSG-code [31370](https://epsg.io/31370)):

```{r}
get_locs(watina,
         bbox = c(xmin = 1.4e+5,
                  xmax = 1.7e+5,
                  ymin = 1.6e+5,
                  ymax = 1.9e+5))
#> # Source:     lazy query [?? x 12]
#> # Database:   Microsoft SQL Server
#> # Ordered by: area_code, loc_code, obswell_rank, area_code, loc_code
#>    loc_wid loc_code area_code area_name      x      y loc_validitycode
#>      <int> <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>           
#>  1   10261 ATGP002  ATG       Antitank… 169572 186043 VLD             
#>  2    5629 DORP016  DOR       Dorent    158521 186074 VLD             
#>  3    5630 DORP017  DOR       Dorent    158442 185925 VLD             
#>  4    5632 DORP019  DOR       Dorent    158304 184915 VLD             
#>  5    5634 DORP020  DOR       Dorent    158352 184259 VLD             
#>  6    5635 DORP021  DOR       Dorent    158103 183853 VLD             
#>  7    5586 DORP022  DOR       Dorent    157715 184013 VLD             
#>  8    5587 DORP023  DOR       Dorent    154284 182533 VLD             
#>  9    6420 DORP024  DOR       Dorent    154823 182431 VLD             
#> 10    5580 DORP027  DOR       Dorent    155151 182190 VLD             
#> # … with more rows, and 5 more variables: loc_validity <chr>,
#> #   loc_typecode <chr>, loc_typename <chr>, filterdepth <dbl>,
#> #   soilsurf_ost <dbl>
```

- if you already have specific location codes in mind, you can provide these directly as a vector:

```{r}
get_locs(watina,
         loc_vec = c("KBRP081", "KBRP090", "KBRP095"))
#> # Source:     lazy query [?? x 12]
#> # Database:   Microsoft SQL Server
#> # Ordered by: area_code, loc_code, obswell_rank, area_code, loc_code
#>   loc_wid loc_code area_code area_name      x      y loc_validitycode
#>     <int> <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>           
#> 1     664 KBRP081  KBR       Potpolde… 145846 204871 VLD             
#> 2     698 KBRP090  KBR       Potpolde… 145988 204702 VLD             
#> 3     709 KBRP095  KBR       Potpolde… 145949 204712 VLD             
#> # … with 5 more variables: loc_validity <chr>, loc_typecode <chr>,
#> #   loc_typename <chr>, filterdepth <dbl>, soilsurf_ost <dbl>
```

- you can use a geospatial filter of class [sf](https://r-spatial.github.io/sf/) to make selections, by using the `mask` argument:

```{r echo=FALSE, include=FALSE}
westfl <- 
    "https://geoservices.informatievlaanderen.be/overdrachtdiensten/VRBG/wfs" %>%
    httr::parse_url() %>%
    purrr::list_merge(query = list(request = "GetFeature",
                                   typeName = "VRBG:Refprv",
                                   cql_filter="NAAM='West-Vlaanderen'",
                                   srsName = "EPSG:31370",
                                   outputFormat = "text/xml; subtype=gml/3.1.1")) %>%
    httr::build_url() %>%
    read_sf(crs = 31370) %>% 
    st_cast("GEOMETRYCOLLECTION")
```


```{r}
get_locs(watina,
         mask = westfl,  # sf polygon of the 'West-Vlaanderen' province
         buffer = 0)
#> As a mask always invokes a collect(), the argument 'collect = FALSE' will be ignored.
#> Warning in get_locs(watina, mask = westfl, buffer = 0): Dropped 40
#> locations from which x or y coordinates were missing.
#> # A tibble: 781 x 11
#>    loc_code area_code area_name     x      y loc_validitycode loc_validity
#>  * <chr>    <chr>     <chr>     <dbl>  <dbl> <chr>            <chr>       
#>  1 ASSP001  ASS       Assebroe… 74324 208838 VLD              Gevalideerd 
#>  2 ASSP002  ASS       Assebroe… 74526 208787 VLD              Gevalideerd 
#>  3 ASSP003  ASS       Assebroe… 74169 208790 VLD              Gevalideerd 
#>  4 ASSP004  ASS       Assebroe… 74108 208605 VLD              Gevalideerd 
#>  5 ASSP005  ASS       Assebroe… 73910 208646 VLD              Gevalideerd 
#>  6 ASSP006  ASS       Assebroe… 74033 208916 VLD              Gevalideerd 
#>  7 ASSP007  ASS       Assebroe… 73609 208954 VLD              Gevalideerd 
#>  8 ASSP008  ASS       Assebroe… 73948 209104 VLD              Gevalideerd 
#>  9 ASSP009  ASS       Assebroe… 73613 208584 VLD              Gevalideerd 
#> 10 ASSP010  ASS       Assebroe… 73583 208390 VLD              Gevalideerd 
#> # … with 771 more rows, and 4 more variables: loc_typecode <chr>,
#> #   loc_typename <chr>, filterdepth <dbl>, soilsurf_ost <dbl>
```

Note that the default value for `buffer` is 10 meters; it is used to enlarge mask (or shrink it, if `buffer` < 0).
Also, note there are a few warnings to inform about dropped locations and duplicated coordinates.

Instead of leaving the output as a `tbl_lazy` object, you can also retrieve it as a tibble (i.e. an easier-to-use type of dataframe) by setting `collect = TRUE`.

```{r}
get_locs(watina,
         area_codes = c("KAL", "KBR"),
         loc_type = c("P", "S"),
         collect = TRUE
         )
#> # A tibble: 223 x 11
#>    loc_code area_code area_name      x      y loc_validitycode loc_validity
#>    <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>            <chr>       
#>  1 KBRP081  KBR       Potpolde… 145846 204871 VLD              Gevalideerd 
#>  2 KBRP090  KBR       Potpolde… 145988 204702 VLD              Gevalideerd 
#>  3 KBRP095  KBR       Potpolde… 145949 204712 VLD              Gevalideerd 
#>  4 KBRS001  KBR       Potpolde… 146937 205586 VLD              Gevalideerd 
#>  5 KALP030  KAL       Kalmthou… 152990 238600 VLD              Gevalideerd 
#>  6 KALP035  KAL       Kalmthou… 153000 236850 VLD              Gevalideerd 
#>  7 KALP039  KAL       Kalmthou… 160940 235060 VLD              Gevalideerd 
#>  8 KALP052  KAL       Kalmthou… 153822 233761 VLD              Gevalideerd 
#>  9 KALP053  KAL       Kalmthou… 153781 233736 VLD              Gevalideerd 
#> 10 KALP059  KAL       Kalmthou… 154754 232459 VLD              Gevalideerd 
#> # … with 213 more rows, and 4 more variables: loc_typecode <chr>,
#> #   loc_typename <chr>, filterdepth <dbl>, soilsurf_ost <dbl>
```

Note that the lazy object contains a `loc_wid` column, which is an internal database variable to identify locations and make relations between tables.
It should not be regarded as a stable ID; hence don't store it for later use (note that lazy objects don't store data).
Consequently, the `loc_wid` column is omitted when using `collect = TRUE`.

Optionally, you can retrieve the individual **observation wells** that are linked to each location (where each observation well is from a different timeframe).
In this case, more attributes are returned (specific observation well attributes).
The below example also demonstrates the use of `tibble::glimpse()` to preview all columns:

```{r}
get_locs(watina,
         obswells = TRUE,
         area_codes = c("KAL", "KBR"),
         loc_type = c("P", "S")) %>% 
  glimpse()
#> # Source:     lazy query [?? x 14]
#> # Database:   Microsoft SQL Server
#> # Ordered by: area_code, loc_code, obswell_rank
#>    loc_wid loc_code area_code area_name      x      y loc_validitycode
#>      <int> <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>           
#>  1    2824 KALP022  KAL       Kalmthou… 155350 230813 VLD             
#>  2    1546 KALP030  KAL       Kalmthou… 152990 238600 VLD             
#>  3   10069 KALP034  KAL       Kalmthou… 152962 231345 VLD             
#>  4    1549 KALP035  KAL       Kalmthou… 153000 236850 VLD             
#>  5    1552 KALP039  KAL       Kalmthou… 160940 235060 VLD             
#>  6    4327 KALP050  KAL       Kalmthou… 153881 233802 VLD             
#>  7    6210 KALP051  KAL       Kalmthou… 153822 233761 VLD             
#>  8    1562 KALP052  KAL       Kalmthou… 153822 233761 VLD             
#>  9    1563 KALP053  KAL       Kalmthou… 153781 233736 VLD             
#> 10    6180 KALP054  KAL       Kalmthou… 154689 232426 VLD             
#> # … with more rows, and 7 more variables: loc_validity <chr>,
#> #   loc_typecode <chr>, loc_typename <chr>, obswell_code <chr>,
#> #   obswell_rank <int>, filterdepth <dbl>, soilsurf_ost <dbl>
```

With `get_locs(obswells = FALSE)`, which is the default, the observation well variables are aggregated for each location using one of four methods.
The method is set with the argument `obswell_aggr`.
The default, `obswell_aggr = "latest"`, returns the attributes of the most recent observation well that fulfills the `filterdepth_range` and `filterdepth_na` criteria.

## Disconnecting

At the end of your work (a script, a bookdown document etc.), close the connection:

```{r}
dbDisconnect(watina)
```



