---
title: "Using XG3 values to characterize and select locations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connecting and retrieving locations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(watina)
library(sf)
library(dplyr)
library(stringr)
library(knitr)
```

## Overview

XG3 is an umbrella for the three geohydrological variables LG3, HG3 and VG3, which are yearly statistics which can be calculated from series of water level measurements (on condition that data are enough frequent).
More explanation on these variables is given by the documentation of `get_xg3()`.

Let's suppose that we want to select locations (within an area, i.e. after having used `get_locs()`) for which:

- multiple successive years are available with both an LG3 and a HG3 value. We call such a multi-year series of an XG3 variable **an XG3 series**.
- we restrict the concept of an XG3 series to the situations where the minimum series length is 5 years, and where years with missing data (for LG3 and/or HG3) must be single (gaps in a series can not be longer than 1 year).
However the user is free to choose this, using the `min_dur` and `max_gap` arguments.
- the last year of the XG3 series must be 2015 or later.

We can store specific conditions in a dataframe; the available statistics are explained by the documentation of `eval_xg3_avail()` and `eval_xg3_series()`:

```{r}
conditions_df <-
  tribble(
  ~xg3_variable, ~statistic, ~criterion, ~direction,
  "combined", "ser_lastyear", 2015, "min"
  )
conditions_df %>% 
  kable
```

The other aspects (which XG3 types?, `max_gap`, `min_dur`) are directly specified in the `selectlocs_xg3()` function:

```{r}
watina <- connect_watina()
get_locs(watina,
         area_codes = "TOR",
         loc_type = c("P", "S")) %>% 
  get_xg3(watina, 1900) %>% 
  selectlocs_xg3(xg3_type = c("L", "H"),
                  max_gap = 1,
                  min_dur = 5,
                  conditions = conditions_df)
```

With the argument `list = TRUE` you can also obtain intermediate test results.
For further information, see the documentation of the `selectlocs_xg3()` function.


## The basics: obtaining XG3 values

The `get_xg3()` function in the above example retrieved XG3 values, from given locations, as a lazy object (unless `collect = TRUE`).
A timeframe (in hydrological years) is used to filter XG3 values in Watina (arguments `startyear` and `endyear`).
By default, the `endyear` argument is set to the previous hydrological year.
Here we only provide the `startyear`:

```{r}
mylocs <- get_locs(watina, area_codes = "KAL")
mylocs %>% get_xg3(watina, 2010)
```

Retrieving the data locally:

```{r}
mylocs %>% get_xg3(watina, 2010, collect = TRUE)
```

By default, the "local" vertical coordinate reference system (CRS) is used to express water levels, but also the CRS Ostend height (EPSG [5710](https://epsg.io/5710), also known as 'TAW' or 'DNG') is available from the database by setting `vert_crs = "ostend"`.

```{r}
mylocs %>% get_xg3(watina, 2010, vert_crs = "ostend")
```

Joining results to `mylocs`:

```{r}
mylocs %>%
  get_xg3(watina, 2010) %>%
  left_join(mylocs %>%
            select(-loc_wid),
            .) %>%
  collect
```


## Processing XG3 values and selecting locations

You can characterize the locations of a dataset of XG3 values, using the `eval_xg3_avail()` and `eval_xg3_series()` functions.
You are invited to read their documentation and try their examples!

The `selectlocs_xg3()` function, of which we saw a demonstration above, sits on top of these.
It calls `eval_xg3_avail()` and `eval_xg3_series()` by itself, but alternatively the user can provide the result of those functions as input to `selectlocs_xg3()`.



```{r include=FALSE}
DBI::dbDisconnect(watina)
```

