---
title: "Using XG3 values to characterize and select locations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using XG3 values to characterize and select locations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(watina)
library(sf)
library(dplyr)
library(stringr)
library(knitr)
```

_General note: the below vignette contains frozen output of 16 Sep 2019._
_This makes it possible to build the package with vignettes without access to the Watina database._

## Overview

XG3 is an umbrella for the three geohydrological variables LG3, HG3 and VG3, which are yearly statistics which can be calculated from series of water level measurements (on condition that data are frequent enough).
More explanation on these variables is given by the documentation of `get_xg3()`.

Let's suppose that we want to select locations (within an area, i.e. after having used `get_locs()`) for which:

- multiple successive years are available with both an LG3 and a HG3 value. We call such a multi-year series of an XG3 variable **an XG3 series**.
- we restrict the concept of an XG3 series to the situations where the minimum series length is 5 years, and where years with missing data (for LG3 and/or HG3) must be single (gaps in a series can not be longer than 1 year).
However the user is free to choose this, using the `min_dur` and `max_gap` arguments.
- the last year of the XG3 series must be 2015 or later.

We can store specific conditions in a dataframe; the available statistics are explained by the documentation of `eval_xg3_avail()` and `eval_xg3_series()`:

```{r eval=TRUE}
conditions_df <-
  tribble(
  ~xg3_variable, ~statistic, ~criterion, ~direction,
  "combined", "ser_lastyear", 2015, "min"
  )
conditions_df %>% 
  kable
```

The other aspects (which XG3 types?, `max_gap`, `min_dur`) are directly specified in the `selectlocs_xg3()` function:

```{r}
watina <- connect_watina()
get_locs(watina,
         area_codes = "TOR",
         loc_type = c("P", "S")) %>% 
  get_xg3(watina, 1900) %>% 
  selectlocs_xg3(xg3_type = c("L", "H"),
                  max_gap = 1,
                  min_dur = 5,
                  conditions = conditions_df)
#> Dropped 19 locations in series evaluation because no XG3 series are available at those locations, for the requested XG3 variables: TORP012, TORP013, TORP019, TORP023, TORP025, TORP026, TORP027, TORP030, TORP031, TORS001, TORS002, TORS003, TORS004, TORS005, TORS006, TORS007, TORS008, TORS010, TORS012
#> # A tibble: 16 x 1
#>    loc_code
#>    <chr>   
#>  1 TORP001 
#>  2 TORP002 
#>  3 TORP004 
#>  4 TORP005 
#>  5 TORP010 
#>  6 TORP011 
#>  7 TORP014 
#>  8 TORP015 
#>  9 TORP016 
#> 10 TORP018 
#> 11 TORP020 
#> 12 TORP022 
#> 13 TORP024 
#> 14 TORP028 
#> 15 TORP029 
#> 16 TORS009
```

With the argument `list = TRUE` you can also obtain intermediate test results.
For further information, see the documentation of the `selectlocs_xg3()` function.


## The basics: obtaining XG3 values

The `get_xg3()` function in the above example retrieved XG3 values, from given locations, as a lazy object (unless `collect = TRUE`).
A timeframe (in hydrological years) is used to filter XG3 values in Watina (arguments `startyear` and `endyear`).
By default, the `endyear` argument is set to the previous hydrological year.
Here we only provide the `startyear`:

```{r}
mylocs <- get_locs(watina, area_codes = "KAL")
mylocs %>% get_xg3(watina, 2010)
#> # Source:     lazy query [?? x 5]
#> # Database:   Microsoft SQL Server
#> # Ordered by: area_code, loc_code, loc_code, hydroyear
#>    loc_code hydroyear lg3_lcl hg3_lcl vg3_lcl
#>    <chr>        <int>   <dbl>   <dbl>   <dbl>
#>  1 KALP022       2014  -0.948  -0.135  -0.473
#>  2 KALP022       2015  -1.05   -0.112  -0.340
#>  3 KALP022       2016  -1.07   -0.124  -0.269
#>  4 KALP022       2017  NA      NA      -0.326
#>  5 KALP079       2010  NA      NA      -0.873
#>  6 KALP079       2011  NA      NA      -0.747
#>  7 KALP079       2012  NA      NA      -0.773
#>  8 KALP079       2013  NA      NA      -0.64 
#>  9 KALP081       2010  NA      NA      -0.557
#> 10 KALP081       2013  NA      NA      -0.413
#> # … with more rows
```

Retrieving the data locally:

```{r}
mylocs %>% get_xg3(watina, 2010, collect = TRUE)
#> # A tibble: 46 x 5
#>    loc_code hydroyear lg3_lcl hg3_lcl vg3_lcl
#>    <chr>        <int>   <dbl>   <dbl>   <dbl>
#>  1 KALP022       2014  -0.948  -0.135  -0.473
#>  2 KALP022       2015  -1.05   -0.112  -0.340
#>  3 KALP022       2016  -1.07   -0.124  -0.269
#>  4 KALP022       2017  NA      NA      -0.326
#>  5 KALP079       2010  NA      NA      -0.873
#>  6 KALP079       2011  NA      NA      -0.747
#>  7 KALP079       2012  NA      NA      -0.773
#>  8 KALP079       2013  NA      NA      -0.64 
#>  9 KALP081       2010  NA      NA      -0.557
#> 10 KALP081       2013  NA      NA      -0.413
#> # … with 36 more rows
```

By default, the "local" vertical coordinate reference system (CRS) is used to express water levels, but also the CRS Ostend height (EPSG [5710](https://epsg.io/5710), also known as 'TAW' or 'DNG') is available from the database by setting `vert_crs = "ostend"`.

```{r}
mylocs %>% get_xg3(watina, 2010, vert_crs = "ostend")
#> # Source:     lazy query [?? x 5]
#> # Database:   Microsoft SQL Server
#> # Ordered by: area_code, loc_code, loc_code, hydroyear
#>    loc_code hydroyear lg3_ost hg3_ost vg3_ost
#>    <chr>        <int>   <dbl>   <dbl>   <dbl>
#>  1 KALP022       2014    22.5    23.4    23.0
#>  2 KALP022       2015    22.4    23.4    23.1
#>  3 KALP022       2016    22.4    23.4    23.2
#>  4 KALP022       2017    NA      NA      23.2
#>  5 KALP079       2010    NA      NA      19.6
#>  6 KALP079       2011    NA      NA      19.7
#>  7 KALP079       2012    NA      NA      19.7
#>  8 KALP079       2013    NA      NA      19.8
#>  9 KALP081       2010    NA      NA      20.9
#> 10 KALP081       2013    NA      NA      21.0
#> # … with more rows
```

Joining results to `mylocs`:

```{r}
mylocs %>%
  get_xg3(watina, 2010) %>%
  left_join(mylocs %>%
            select(-loc_wid),
            .) %>%
  collect
#> Joining, by = "loc_code"
#> # A tibble: 109 x 13
#>    loc_code area_code area_name      x      y loc_validitycode loc_validity
#>    <chr>    <chr>     <chr>      <dbl>  <dbl> <chr>            <chr>       
#>  1 KALP030  KAL       Kalmthou… 152990 238600 VLD              Gevalideerd 
#>  2 KALP035  KAL       Kalmthou… 153000 236850 VLD              Gevalideerd 
#>  3 KALP039  KAL       Kalmthou… 160940 235060 VLD              Gevalideerd 
#>  4 KALP052  KAL       Kalmthou… 153822 233761 VLD              Gevalideerd 
#>  5 KALP053  KAL       Kalmthou… 153781 233736 VLD              Gevalideerd 
#>  6 KALP059  KAL       Kalmthou… 154754 232459 VLD              Gevalideerd 
#>  7 KALP060  KAL       Kalmthou… 154702 232476 VLD              Gevalideerd 
#>  8 KALP061  KAL       Kalmthou… 154559 230477 VLD              Gevalideerd 
#>  9 KALP062  KAL       Kalmthou… 154553 230423 VLD              Gevalideerd 
#> 10 KALP063  KAL       Kalmthou… 154669 230340 VLD              Gevalideerd 
#> # … with 99 more rows, and 6 more variables: loc_typecode <chr>,
#> #   loc_typename <chr>, hydroyear <int>, lg3_lcl <dbl>, hg3_lcl <dbl>,
#> #   vg3_lcl <dbl>
```


## Processing XG3 values and selecting locations

You can characterize the locations of a dataset of XG3 values, using the `eval_xg3_avail()` and `eval_xg3_series()` functions.
You are invited to read their documentation and try their examples!

The `selectlocs_xg3()` function, of which we saw a demonstration above, sits on top of these.
It calls `eval_xg3_avail()` and `eval_xg3_series()` by itself, but alternatively the user can provide the result of those functions as input to `selectlocs_xg3()`.



```{r include=FALSE}
DBI::dbDisconnect(watina)
```

