---
title: "Using hydrochemical data to characterize and select locations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using hydrochemical data to characterize and select locations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(watina)
library(sf)
library(dplyr)
library(stringr)
library(knitr)
```

## Overview

Let's have a look at how chemical data, retrieved from Watina, look like.
The below example requests available hydrochemical data since 2010 from locations in the area 'Zwarte Beek':

```{r}
watina <- connect_watina()
mydata <-
    get_locs(watina, area_codes = "ZWA") %>% 
    get_chem(watina, "1/1/2010")
mydata %>% 
  count
class(mydata)
mydata %>% 
  head(15) %>% 
  collect %>% 
  kable
```

The first and last year of the dataset are:

```{r}
mydata %>%
    pull(date) %>%
    lubridate::year(.) %>%
    (function(x) c(firstyear = min(x), lastyear = max(x)))
```

Let's suppose that we now want to select locations for which:

- both N-NO3 (nitrate nitrogen) and P-PO4 (orthophosphate phosphorus) have at least two measurement dates;
- the first date of P-PO4 is not later than 1/1/2014;
- the first and last year with P-PO4 data are at least 4 years apart, i.e. they span at least 5 calendar years

To express criteria, we need the numerical value of a date:

```{r}
as.numeric(lubridate::dmy("1/1/2014"))
```

We can store specific conditions in a dataframe; the available statistics are explained by the documentation of `eval_chem()`:

```{r}
conditions_df <-
    tribble(
        ~chem_variable, ~statistic, ~criterion, ~direction,
        "N-NO3", "nrdates", 2, "min",
        "P-PO4", "nrdates", 2, "min",
        "P-PO4", "firstdate", 16071, "max",
        "P-PO4", "timespan_years", 5, "min"
    )
conditions_df %>% 
  kable
```

You can also separately limit the chemical variables to be evaluated with the `chem_var` argument in the `selectlocs_chem()` function:

```{r}
myresult <-
    mydata %>%
    selectlocs_chem(data_type = "data",
                    chem_var = c("N-NO3", "P-PO4"),
                    conditions = conditions_df)
myresult
```

With the argument `list = TRUE` you can also obtain intermediate test results.
For further information, see the documentation of the `selectlocs_chem()` function.


## The basics: obtaining hydrochemical data

The `get_chem()` function in the above example retrieved hydrochemical data, from given locations, as a lazy object (unless `collect = TRUE`).
A timeframe is used to filter hydrochemical data in Watina (with arguments `startdate` and `enddate`).
By default, the `enddate` argument is set as _today_.
Here we only provide the `startdate`:

```{r}
mylocs <- get_locs(watina, area_codes = "ZWA")
mylocs %>% get_chem(watina, "1/1/2017")
```

Retrieving the data locally:

```{r}
mylocs %>% get_chem(watina, "1/1/2017", collect = TRUE)
```

Note below, the different number of results when using non-default settings for filtering according to _electroneutrality_!
The default range of electroneutrality is from -0.1 to 0.1 (argument `en_range`).

```{r}
mylocs %>% get_chem(watina, "1/1/2017") %>% count
mylocs %>% get_chem(watina, "1/1/2017", strict_en = TRUE) %>% count
mylocs %>% get_chem(watina, "1/1/2017", en_range = c(-1, 1)) %>% count
```

With `strict_en = TRUE`, samples for which no electroneutrality could be calculated are discarded!

Equivalence concentrations instead of mass concentrations can be returned with `conc_type = "eq"`:

```{r}
mylocs %>% get_chem(watina, "1/1/2017", conc_type = "eq")
```

Joining results to `mylocs`:

```{r}
mylocs %>%
get_chem(watina, "1/1/2017") %>%
    left_join(mylocs %>%
                  select(-loc_wid),
              .) %>%
    collect
```


## Processing hydrochemical data and selecting locations

You can characterize the locations of a dataset of hydrochemical data, using the `eval_chem()` function.
You are invited to read its documentation and try its examples!

The `selectlocs_chem()` function, of which we saw a demonstration above, calls `eval_chem()` by itself.
Alternatively the user can provide the result of those functions as input to `selectlocs_chem()`.

```{r include=FALSE}
DBI::dbDisconnect(watina)
```

