---
title: "Creating plots based on hydrochemical data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating plots based on hydrochemical data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)

library(watina)
library(dplyr)
library(knitr)
library(ggplot2)
```

_General note: the below vignette contains frozen output of 7 Nov 2025._
_This makes it possible to build the package with vignettes without access to the Watina database._

## Retrieve chemical data from Watina

We use the `get_chem()` function to retrieve hydrochemical data.
The example requests available hydrochemical data since 2024 from locations 
in the area 'Kesterbeek' and saves it locally `(collect = TRUE)`: 

```{r, eval = FALSE}
watina <- connect_watina()
my_chem_data <-
    get_locs(watina, area_codes = "KES") %>% 
    get_chem(watina, "1/1/2024", collect = TRUE)

```

```{r, eval = FALSE, include = FALSE}
dbDisconnect(watina)
```

```{r, eval = FALSE}
my_chem_data
#> # A tibble: 272 x 10
#>    loc_code date       lab_project_id lab_sample_id chem_variable  value
#>    <chr>    <date>     <chr>          <chr>         <chr>          <dbl>
#>  1 KESP001  2024-07-04 0              42052         Al             0.050
#>  2 KESP001  2024-07-04 0              42052         Ca             109.881
#>  3 KESP001  2024-07-04 0              42052         Cl             11.349 
#>  4 KESP001  2024-07-04 0              42052         CondF          599.000   
#>  5 KESP001  2024-07-04 0              42052         CondL          606.700  
#>  6 KESP001  2024-07-04 0              42052         Fe             2.788
#>  7 KESP001  2024-07-04 0              42052         HCO3           375.292
#>  8 KESP001  2024-07-04 0              42052         K              0.970
#>  9 KESP001  2024-07-04 0              42052         Mg             8.960
#> 10 KESP001  2024-07-04 0              42052         Mn             0.792
#> # … with 262 more rows, and 4 more variables: unit <chr>,
#> #   below_loq <lgl>, loq <dbl>, elneutr <dbl>
```
```{r, include = FALSE}
# needed if we build the package with vignettes without access to the Watina database
my_chem_data <- data.frame(
    loc_code = rep(c("KESP001", "KESP003"), each = 6),
    date = as.Date(rep(c("2024-07-04", "2024-11-13"), each = 3, times = 2)),
    lab_project_id = rep(0, 12),
    lab_sample_id = rep(c("42052", "42307", "42048", "42303"), each = 3),
    elneutr = c(0.005029, 0.005029, 0.005029, 0.014906, 0.014906, 0.014906, 
                0.041433, 0.041433, 0.041433, 0.024119, 0.024119, 0.024119),
    chem_variable = rep(c("Ca", "Cl", "CondL"), 4),
    value = c(109.881, 11.349, 606.7, 110.967, 8.763, 579.6,
              135.331, 31.653, 730.6, 124.289, 31.781, 677.2),
    unit = rep(c("mg/l", "mg/l", "µS/cm"), 4),
    below_loq = rep(FALSE, 12),
    loq = rep(-99, 12)
    )
```

We will use this dataset to demonstrate the plotting options available in the 
Watina package.

## Plot the ionic ratio against the electric conductivity (Van Wirdum)

This diagram, devised by Geert Van Wirdum, shows the chemical similarity of a 
water sample to references samples such as atmocline water (rainwater), 
thalassocline water (seawater) and lithocline water (calcium rich, fresh 
groundwater). The diagram plots the ionic ratio, 
that is *[Ca2+] / ([Ca2+] + [Cl-])*, against the logarithm of the electric 
conductivity at 25°C.

The electric conductivity is available for most samples in Watina, but we need 
to compute the ionic ratio. 
We could do this manually for each sample but the `watina` package also contains 
a `calc_ir()` function that can be applied directly to the output of the 
`get_chem()` function to calculate the ionic ratio:

```{r}
my_chem_data_vanwirdum <- calc_ir(my_chem_data)
```

The obtained dataset is a broad table with the following fields:

```{r }
names(my_chem_data_vanwirdum)
```

The new fields `Ca_meq` and `Cl_meq` give the Ca and Cl concentrations in 
milliequivalents and the field `ir` contains the calculated ionic ratio.

Now we can create the diagram of the ionic ratio against the (log of) 
electric conductivity at 25°C (the so called Van Wirdum diagram) with 
the `ggplot_vanwirdum_background()` function. 
This function will create the background of a Van Wirdum diagram, 
with reference points for lithotrophic water (LI), atmotrophic water (AT), 
thalassotrophic water (TH) and optionally molunotrophic water (RH - polluted 
water as found in the Rhine).
 
```{r , fig.height=3.5, fig.width=6}
# background with standard options:
print(ggplot_vanwirdum_background() )
```

We can show the mixing contours of the reference points with a curve instead of a line
and plot the reference point for polluted river water using respectively the 
`contour` and `rhine` arguments:

```{r, fig.height=3.5, fig.width=6}
print(ggplot_vanwirdum_background(contour = "curve",
                                  rhine = TRUE))
```


Now let's add our own data for the valley of the Kesterbeek.
In this example we plot 2 locations that were sampled twice in 2024:

```{r, fig.height=3.5, fig.width=6}
# add your own data with EC as x and IR as y and format as you wish
ggplot_vanwirdum_background(contour = "curve",
             lang = "en",
             rhine = TRUE) +
  geom_point(data = my_chem_data_vanwirdum %>% head(n = 4),
             aes(x = CondL, y = ir, 
                 colour = loc_code,
                 shape = as.factor(date)), size = 3) +
    scale_colour_manual(name = "Location", values = c("dodgerblue", "red")) + 
    scale_shape_discrete(name = "Date")
```


All the usual options of `ggplot2` are available to customize the layout of the plot.




